---
title: "Analysis-of-Suicide-Trends-Globally-in-R"
Script Owner: "Abhishek Dubey"
output: html_notebook
---

Installing Packages 
```{r}
install.packages("ggalt")
install.packages("countrycode")
install.packages("rworldmap")
install.packages("gridExtra")
install.packages("broom")
install.packages("magrittr")

```


```{r}
library(tidyverse)
library(ggalt)
library(countrycode)
library(rworldmap)
library(gridExtra)
library(broom)
library(readr)
library(magrittr)
```
Data Import
```{r}

 Country_df<- read_csv("https://raw.githubusercontent.com/Abhidubey96/Analysis-of-Suicide-Trends-Globally-in-R/main/Country%20Data.csv")

 Suicide_df <- read_csv("https://raw.githubusercontent.com/Abhidubey96/Analysis-of-Suicide-Trends-Globally-in-R/main/Suicide%20Data%20by%20Country.csv")

```
View Data
```{r}

View(Country_df)
View(Suicide_df)

```
Merging Both 2 files
```{r}
df <- cbind(Suicide_df, Country_df)

View(df)
```

CLeaning and Wrangling of data
```{r}
summary(df)

str(df)

# Removing unnecessary columns 
# Renaming the variable names 
df <- df %>% 
  select(-c(`HDI for year`, `suicides/100k pop`)) %>%
  rename(gdp_per_year = `gdp_for_year ($)`, 
         gdp_per_capita = `gdp_per_capita ($)`, 
         country_year = `country-year`) %>%
  as.data.frame()

View(df)
str(df)

```
Extensive Cleaning and Wrangling in Data
```{r}
# data is missing for 2016 year for many countries
# every country should have 12 observations based upon the age and gender. But Data for 2016 shows less entries. Also some countries are missing totally.
# So its better to remove it from data frame df
df <- df %>%
  filter(year != 2016) %>% 
  select(-country_year)

# So now we have only data from 1985 to 2015
View(df)

# Making Continent variable by using country code library
df$continent <- countrycode(sourcevar = df[, "country"],
                              origin = "country.name",
                              destination = "continent")

View(df)

#Changing Data type for variables age and generation and making them ordinal by using factor function

df$age <- factor(df$age, ordered = T, 
                   levels = c("5-14 years",
                              "15-24 years", 
                              "25-34 years", 
                              "35-54 years", 
                              "55-74 years", 
                              "75+ years"))

df$generation <- factor(df$generation, 
                   ordered = T, 
                   levels = c("G.I. Generation", 
                              "Silent",
                              "Boomers", 
                              "Generation X", 
                              "Millenials", 
                              "Generation Z"))

# Global suicide rate over time 
Global <- (sum(as.numeric(df$suicides_no)) / sum(as.numeric(df$population))) * 100000

View(df)

summary(df)
glimpse(df)

# Now our Data Frame is cleaned and ready for visualizations
```
Visualization 1
Heat Map of suicide cases globally

Iteration 1
This cell will lot of time because suicides no entries are more, better to comment this cell.
```{r}

mapdata <- joinCountryData2Map(df, joinCode = "NAME", nameJoinColumn = "country")

mapCountryData(mapdata, 
nameColumnToPlot="suicides_no",  
colourPalette = "heat",
mapTitle="Total suicide cases globally from 1985 to 2015",
mapRegion = "world",
oceanCol="deepskyblue",
catMethod = "pretty")
```
Iteration 2
Adding suicide per 100k population

```{r}

region <- df %>%
  group_by(country) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

countrydata1 <- joinCountryData2Map(region, joinCode = "NAME", nameJoinColumn = "country")

par(mar=c(0, 0, 0, 0)) # margins


mapCountryData(countrydata1, 
nameColumnToPlot="suicide_per_100k",  
colourPalette = "heat",
mapTitle="Total suicide cases globally as per (suicide per 100k population). Shows overall cases by country from 1985 to 2015",
mapRegion = "world",
oceanCol="deepskyblue", 
missingCountryCol="gray",
borderCol = "black",
catMethod = "pretty")
```
Visualizing suicide cases for america and europe
```{r}


region <- df %>%
  group_by(country) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

countrydata <- joinCountryData2Map(region, joinCode = "NAME", nameJoinColumn = "country")

par(mar=c(0, 0, 0, 0)) # margins
mapCountryData(countrydata, 
nameColumnToPlot="suicide_per_100k", 
mapTitle="Suicide cases in europe",
mapRegion = "europe",
oceanCol="deepskyblue", 
missingCountryCol="gray",
borderCol = "black",
catMethod = "pretty")



mapCountryData(countrydata, 
nameColumnToPlot="suicide_per_100k", 
mapTitle="Suicide cases in north america",
mapRegion = "north america",
oceanCol="deepskyblue", 
missingCountryCol="gray",
borderCol = "black",
catMethod = "pretty")
```


Visualization 2

What is the relation between age and generation over years from 1985 to 2015 suicide cases per 100k population?
Is it True that new generation suicide cases globally are growing rapidly?

To Answer above question, we will follow below iterations

visualization for age varibale
```{r}
df %>%
  group_by(age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = age, y = suicide_per_100k, fill = age)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Global suicides per 100k, by Age",
       x = "Age", 
       y = "Suicides per 100k") +
  theme(legend.position = "none") + 
  scale_y_continuous(breaks = seq(0, 30, 1), minor_breaks = F)
```
visualization of generation variable
```{r}
df %>%
  group_by(generation) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = generation, y = suicide_per_100k, fill = generation)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Global suicides per 100k, by generation",
       x = "generation", 
       y = "Suicides per 100k") +
  theme(legend.position = "none") + 
  scale_y_continuous(breaks = seq(0, 30, 1), minor_breaks = F)
```

final visualization by merging 1st and 2 nd visualization of age and generation. plotting them with year ( over time )
```{r}

df %>%
  group_by(generation, age, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = factor(generation, ordered = F))) + 
  geom_point() + 
  geom_line() + 
  facet_grid(age ~ ., scales = "free_y") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = NULL) +
  labs(title = "Relationship between Generations, Age per group & Year from 1985 to 2015", 
       x = "Year", 
       y = "Suicides per 100k", 
       col = "Generation") + 
  theme(legend.position = "bottom")
```



Visualization 3

What is the global trend of suicide cases per 100k population from 1985 to 2015?

If we talk about some specific countries from Europe and North America what’s the trend?

Iteration 1

Analyzing countries with high gdp per capita income

```{r}
library(sqldf)

gdp_df <-sqldf("select country, sum(gdp_per_capita) as gdp from df group by country")
gdp_df

# finding 5 countries with max gdp per capita income

gdp_df1 <-sqldf("select country, gdp from gdp_df where gdp > 14300000")
gdp_df1

```
Plotting
```{r}
ggplot(data=gdp_df1, aes(x= country, y=gdp)) +
geom_bar(stat="identity")
```



Iteration 2

Checking Visualization suicide trend over time
For all Countries 
```{r}
df %>%
  group_by(year) %>%
  summarize(population = sum(population), 
            suicides = sum(suicides_no), 
            suicides_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = year, y = suicides_per_100k)) + 
  geom_line(col = "red", size = 1) + 
  geom_point(col = "red", size = 2) + 
  labs(title = "Global Suicides (per 100k)",
       x = "Year", 
       y = "Suicides per 100k") + 
  scale_x_continuous(breaks = seq(1985, 2015, 2)) + 
  scale_y_continuous(breaks = seq(10, 20))
```

Iteration 3

HIGH GDP countries trend of suicide cases over time 
Checking suicide cases for specific countries

```{r}

df_specific <- df %>%
  filter(country %in% c("Iceland", 
                        "Luxembourg",
                        "Norway", 
                        "Switzerland", 
                        "United States"))

df_specific %>% 
  group_by(country, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = country)) + 
  geom_smooth(se = F, span = 0.2) +
  scale_x_continuous(breaks = seq(1985, 2015, 2), minor_breaks = F) + 
  labs(title = " Suicide cases for United States, Iceland, Luxembourg, Switzerland, Norway", 
       subtitle = "Suicides per 100k population globally, 1985 - 2015", 
       x = "Year ", 
       y = "Suicides per 100k Globally", 
       col = "Country")
```



